Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2024-02-20 Manuscript/Manuscript.Rmd
eLife revised 1

```{r echo=FALSE,fig.cap="**Figure 1. Identification of *NOS*-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Volume rendering of the four INNOS neurons with incoming and outgoing synapses. **(E)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Expression of a *NOS* reporter (NOSp::palmi-3xHA; magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub; green) highlights cPRC cilia and the neuropil. **(G)** Immunostaining for NOS (magenta), co-stained for acetylated α-tubulin (acTub; green). **(H)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses."}

@@

The expression of *NOS* in the INNOS interneurons in the cPRC circuit suggests that NO signalling may be involved in UV/violet-avoidance.
To test this, first we asked whether NO is produced during UV/violet stimulation of the larvae.
We injected the fluorescent NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while exposing the region of cPRC cilia to 405 nm violet light (Figure 2---figure supplement 1).
To mark cell outlines, we coinjected mRNA encoding a red-fluorescent reporter (RGECO), allowing the identification of the cPRCs.
Following light stimulation in the region of the ramified cilia of the cPRCs, we detected an increase in DAF-FM fluorescence in the anterior neurosecretory neuropil, the region of INNOS projections.
This increase did not occur in larvae in which we illuminated a control area (Figure 2).

```{r echo=FALSE,fig.cap="**Figure 2. NO produced by UV/violet stimulation to cPRCs.** (**A**) DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. The circles indicate the location of cPRC and control stimulation. The cPRCs are marked by thin lines. (**B**) DAF-FM fluorescence before and during 405 nm light stimulation. (**C**) Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs or a control area (ctr stim.). The purple box indicates the duration of 405 nm stimulation. Individual normalized traces (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation. <br> Figure 2 -- source data 1. DAF-FM fluorescence reads."}

@@

We next tested whether NO signalling regulates UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines using the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early stop codon and thus likely representing null alleles (Figure 3---figure supplement 1A).
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene in *Platynereis* (Figure 3---figure supplement 1B, C).
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light (Figure 3A and Figure 3---figure supplement 2A, B).
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a significantly diminished UV-avoidance response (Figure 3B, C and Figure 3---figure supplement 1C, D).
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae [@Veraszto2018] and reveals a requirement for NOS in UV-avoidance behaviour.
Wild type but not mutant larvae also showed a significantly increase in swimming speed with an increase in ciliary beat frequency under UV light that may be due to downward swimming trajectories (swimming in the direction of gravity) (Figure 3---figure supplement 3).
We also tested directional phototaxis, by exposing larvae to 480 nm directional collimated light from the top of the column (Figure 3---figure supplement 2A).
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] (Figure 3D and Figure 3---figure supplement 1E).

To distinguish between an acute and developmental function of NOS in light responses, we also tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance (Figure 3E, F).
In contrast, phototaxis was not affected (Figure 3G).
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the function of NO signalling in *Drosophila* eye development [@Gibbs1998a].

```{r echo=FALSE,fig.cap="**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.**  (**A**) Swimming trajectories of wild type (WT, n=32) and *NOS* mutant (*NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec before the onset of 395 nm stimulation from the side. (**B**) Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**C**) Comparison of vertical position of wild type and mutant three-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. The wild type (WT, n=32) and *NOS* mutant (*NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) larvae. (**D**) Vertical displacement in 30 sec bins of wild type (10 batches) and mutant (*NOSΔ11/Δ11*, 5 batches and *NOSΔ23/Δ23*, 7 batches) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. (**E**) Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**F**) Comparison of vertical position of control (n=6) and L-NAME-treated (0.1 and 1 mM, n=6 each) three-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. (**G**) Vertical displacement in 30 sec bins of control (15 batches) and L-NAME-treated (0.1 and 1 mM, 15 batches each) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. <br> Figure 3 -- source data 1-7. Source data for panels A-G."}

@@

@@

```{r echo=FALSE,fig.cap="**Figure 2 -- figure supplement 1. Schematic diagrams of the UV stimulation experiment.** (**A**) Left: The larvae were embedded in 2% agarose seawater with the anterior side down. Right: Changes in DAF-FM fluorescence intensities were examined when the ciliary region of the cPRC was stimulated with a 405 nm laser."}
knitr::include_graphics(path = "figures/Fig2_sup1.png")

@@

```{r echo=FALSE,fig.cap="**Figure 3 -- figure supplement 2. Generation and behavioural characterisation of *NOS* CRISPR knockouts.** (**A**) Schematic diagram of the set-up of the behavioural experiment. (**B**) Swimming trajectories of wild type (WT, n=37) and *NOS* mutant (NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) two-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec before the onset of 395 nm stimulation from the side. (**C**)  Vertical displacement in 30 sec bins of wild type (WT, n=37) and mutant (NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) two-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. (**D**) Comparison of vertical position of wild type and mutant two-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. The wild type (WT, n=52) and *NOS* mutant (*NOSΔ11/Δ11*, n=30 and *NOSΔ23/Δ23*, n=16) larvae. (**E**) Vertical position of batches of wild type (12 batches) and mutant (NOSΔ11/Δ11, 8 batches and NOSΔ23/Δ23, 3 batches) two-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. <br> Figure 3 -- figure suppmenet 2 -- source data 1-4. Source data for panels B-E."}
knitr::include_graphics(path = "figures/Fig3_sup2.png")

>>>>

```{r echo=FALSE,fig.cap="**Figure 1. Identification of *NOS*-expressing interneurons (INNOS) within the cPRC circuit.** (**A**) Scanning electron microscopy image of a three-day-old *Platynereis* larva. (**B, C**) Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. (**D**) Volume rendering of the four INNOS neurons with incoming and outgoing synapses. (**E**) Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. (**F**) Expression of a *NOS* reporter (NOSp::palmi-3xHA; magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub; green) highlights cPRC cilia and the neuropil. (**G**) Immunostaining for NOS (magenta), co-stained for acetylated α-tubulin (acTub; green). (**H**) Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses."}

@@

The expression of *NOS* in the INNOS interneurons in the cPRC circuit suggests that NO signalling may be involved in UV/violet-avoidance.
To test this, first we asked whether NO is produced during UV/violet stimulation of the larvae.
We injected the fluorescent NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while exposing the region of cPRC cilia to 405 nm violet light (Figure 2A, B).
To mark cell outlines, we coinjected mRNA encoding a red-fluorescent reporter (RGECO), allowing the identification of the cPRCs.
Following light stimulation in the region of the ramified cilia of the cPRCs, we detected an increase in DAF-FM fluorescence in the anterior neurosecretory neuropil, the region of INNOS projections (Figure 2C).
This increase did not occur in larvae in which we illuminated a control area (Figure 2D).

```{r echo=FALSE,fig.cap="**Figure 2. NO produced by UV/violet stimulation to cPRCs.** (**A**) Left: The larvae were embedded in 2% agarose seawater with the anterior side down. Right: Changes in DAF-FM fluorescence intensities were examined when the ciliary region of the cPRC was stimulated with a 405 nm laser.(**B**) DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. The circles indicate the location of cPRC and control stimulation. The cPRCs are marked by thin lines. (**C**) DAF-FM fluorescence before and during 405 nm light stimulation. (**D**) Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs or a control area (ctr stim.). The purple box indicates the duration of 405 nm stimulation. Individual normalized traces (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation. <br> Figure 2 -- source data 1. DAF-FM fluorescence reads."}

@@

We next tested whether NO signalling regulates UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines using the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early stop codon and thus likely representing null alleles (Figure 3---figure supplement 1A).
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene in *Platynereis* (Figure 3---figure supplement 1B, C).

To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light (Figure 3A and Figure 3---figure supplement 2A, B).
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a significantly diminished UV-avoidance response (Figure 3B, C and Figure 3---figure supplement 1C, D).
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae [@Veraszto2018] and reveals a requirement for NOS in UV-avoidance behaviour.
Wild type but not mutant larvae also showed a significantly increase in swimming speed with an increase in ciliary beat frequency under UV light that may be due to downward swimming trajectories (swimming in the direction of gravity) (Figure 3---figure supplement 3).
We also tested directional phototaxis, by exposing larvae to 480 nm directional collimated light from the top of the column (Figure 3---figure supplement 2A).
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] (Figure 3D and Figure 3---figure supplement 1E).

To distinguish between an acute and developmental function of NOS in light responses, we also tested larvae exposed to the NOS inhibitor L-NAME (Figure 3F).
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance (Figure 3G, H).
In contrast, phototaxis was not affected (Figure 3F).
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the function of NO signalling in *Drosophila* eye development [@Gibbs1998a].

```{r echo=FALSE,fig.cap="**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.** (**A**) Schematic diagram of the set-up of the behavioural experiment. (**B**) Vertical displacement in 30 sec bins of wild type (10 batches) and mutant (*NOSΔ11/Δ11*, 5 batches and *NOSΔ23/Δ23*, 7 batches) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05, **P < 0.01. (**C**) Swimming trajectories of wild type (WT, n=32) and *NOS* mutant (*NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec before the onset of 395 nm stimulation from the side. (**D**) Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**E**) Comparison of vertical position of wild type and mutant three-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. The wild type (WT, n=32) and *NOS* mutant (*NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) larvae. (**F**) Vertical displacement in 30 sec bins of control (15 batches) and L-NAME-treated (0.1 and 1 mM, 15 batches each) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. (**G**) Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**H**) Comparison of vertical position of control (n=6) and L-NAME-treated (0.1 and 1 mM, n=6 each) three-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown.  <br> Figure 3 -- source data 1-7. Source data for panels A-G."}

@@

The TrackMate7 plugin [@Ershov2022] in Fiji ImageJ was used to trace the swimming trajectory of the larvae, using the DoG detector for 3 pixel size and the LAP Tracker, filtered at linking max distance: 10.0 pixels, gap closing max distance: 5.0 pixels.

@@

```{r echo=FALSE,fig.cap="**Figure 2 -- figure supplement 1. "}
knitr::include_graphics(path = "figures/Fig2_sup1.png")

@@

```{r echo=FALSE,fig.cap="**Figure 3 -- figure supplement 2. Generation and behavioural characterisation of *NOS* CRISPR knockouts.** (**B**) Swimming trajectories of wild type (WT, n=37) and *NOS* mutant (NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) two-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec before the onset of 395 nm stimulation from the side. (**C**)  Vertical displacement in 30 sec bins of wild type (WT, n=37) and mutant (NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) two-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. (**D**) Comparison of vertical position of wild type and mutant two-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. The wild type (WT, n=52) and *NOS* mutant (*NOSΔ11/Δ11*, n=30 and *NOSΔ23/Δ23*, n=16) larvae. (**E**) Vertical position of batches of wild type (12 batches) and mutant (NOSΔ11/Δ11, 8 batches and NOSΔ23/Δ23, 3 batches) two-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. <br> Figure 3 -- figure suppmenet 2 -- source data 1-4. Source data for panels B-E."}
knitr::include_graphics(path = "figures/Fig3_sup2.png")
